---
title: "Comulsary Assignment 1"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "2022-09-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 3 (Transformations)


```{r Packages, echo=FALSE}

# Loading necessary packages
library('readr')
library('dplyr')
library('ggplot2')
library('tidyverse')
library('lubridate')
library('zoo')
library('reshape2')
library('factoextra')

```

### Task A

```{r Data, echo=FALSE}

# Loading data
covid <- read_csv("data/covid.csv")
# View(covid)
# str(covid)

# Selecting time-window 
covid <- covid[covid$date >= "2020-03-16" & covid$date <= "2022-01-01", ] 

# Selecting countries
countries <- c('SWE', 'DNK', 'NOR', 'GBR', 'ITA', 'IND')
covid <- covid[covid$iso_code %in% countries,]

# Selecting columns
covid <- select(covid, date, iso_code, new_cases_per_million)
summary(covid)

# Checking that date-column is in correct format
class(covid$date)

```


```{r Data.frame, echo=FALSE}

# Reshaping data.frame
date <- covid$date
date <- date[!duplicated(date)]

swe <- covid[covid$iso_code == 'SWE', ]
swe <- select(swe, new_cases_per_million)
dnk <- covid[covid$iso_code == 'DNK', ]
dnk <- select(dnk, new_cases_per_million)
nor <- covid[covid$iso_code == 'NOR', ]
nor <- select(nor, new_cases_per_million)
gbr <- covid[covid$iso_code == 'GBR', ]
gbr <- select(gbr, new_cases_per_million)
ita <- covid[covid$iso_code == 'ITA', ]
ita <- select(ita, new_cases_per_million)
ind <- covid[covid$iso_code == 'IND', ]
ind <- select(ind, new_cases_per_million)

df <- data.frame(date, swe, dnk, nor, gbr, ita, ind)

# Renaming data.frame
names(df) <- c('date', 'sweden', 'denmark', 'norway', 'britain', 'italia', 'india')

```


### Exploratory analysis

#### Dimensionality and time-index

After restricting the dataset, the series is a univariate time series with 
only one variable that is varying over time and a single object measured. 
The time domain is selected to be from 16.03.2020 to 21.02.2022.
There is one measurement each day, so the resolution is daily 
and the time step is one day. 


```{r Summary statistics}

# Summary statistics
summary(df)

# Missing values
which(is.na(df))

# Creating columns for year, month and day
df <- df %>% mutate(year = lubridate::year(df$date), 
                    month = lubridate::month(df$date), 
                    day = lubridate::day(df$date))

```


```{r Distribution}

# Boxplot of distribution for each country
boxdf = select(df, sweden, denmark, norway, britain, italia, india)
stacked_df <- stack(boxdf)
boxplot(stacked_df$values ~ stacked_df$ind,
        col = rainbow(ncol(df)), xlab='', ylab='New cases')

```

```{r Line plot}

# Reshaping the original data.frame
covid_cases <- melt(covid, id.vars=c('iso_code', 'date'),
                    value.name='Value')

# Line plot containing all the countries
covid_plot <- ggplot(data=covid_cases, aes(x=date,
                                           y=Value, 
                                           group=iso_code,
                                           colour=iso_code)) +
  geom_line() +
  geom_point(size=0.5) + 
  labs(y='New cases', x='')
covid_plot + ggtitle('Daily cases for each country')


```


#### Comments to Task A

The time-series contains no missing values for either of the countries. 
Sweden have many measurements that are 0, due to the fact that they did
not test every day. Denmark, Great Britain and India have some entries that are
below 0. This can be seen from the minimum value in the summary statistics and
also in the line plot.



### Task B

```{r Correlation}

# Pairwise correlation
dfCorr <- select(df, sweden, denmark, norway, italia, india)
corr <- cor(dfCorr, method='pearson')
round(corr, 3)
heatmap(corr)


```


From the correlation matrix it can be seen that Denmark, Norway, Great Britain 
and Italia show high, positive correlation with each other. The value for the 
correlation coefficient is above 0.6, except for the coefficient 
between Norway and Italia, that has a value of 0.429. Sweden show relatively 
low correlation with the other countries. India is negatively correlated 
with the other countries, except from Sweden.

```{r Auto-correlation}

# Producing plots for each of the countries
acf(df$sweden)
# acf(df$denmark)
# acf(df$norway)
# acf(df$britain)
# acf(df$italia)
# acf(df$india)

```


Looking at the auto-correlation for each country, Denmark, Norway, Great Britain
and India shows a slowly decreasing correlation as the lag increases.
All the mentioned countries show some cyclical behavior in the correlation,
with the differences between peaks of 7 days. 
India only shows a slowly decreasing correlation as the lag increases, but 
over the line for significant correlation for all the lags visualized. 
Sweden also show 7-days cycles with highest correlation at the peaks, 
but with much lower correlation for the other days. 
Sweden also show negative values for the correlation coefficient. 


```{r Cross-correlation}

# Producing plots between the different countries

"
# Sweden
ccf(df$sweden, df$denmark)
ccf(df$sweden, df$norway)
ccf(df$sweden, df$britain)
ccf(df$sweden, df$italia)
ccf(df$sweden, df$india)

# Denmark
ccf(df$denmark, df$norway)
ccf(df$denmark, df$britain)
ccf(df$denmark, df$italia)
ccf(df$denmark, df$india)

# Norway
ccf(df$norway, df$denmark)
ccf(df$norway, df$britain)
ccf(df$norway, df$italia)
ccf(df$norway, df$india)

# Britain
ccf(df$britain, df$denmark)
ccf(df$britain, df$norway)
ccf(df$britain, df$italia)
ccf(df$britain, df$india)

# Italia
ccf(df$italia, df$denmark)
ccf(df$italia, df$norway)
ccf(df$italia, df$britain)

# India
ccf(df$india, df$denmark)
ccf(df$india, df$norway)
ccf(df$india, df$britain)
" 
ccf(df$india, df$italia)

```


Denmark, Norway, Great Britain and Italia show statistically significant
positive correlation for the entire period visualized; -25 to +25 days lag.
Sweden show cyclical behavior in the cross-correlation with the other 
countries, with variation in the coefficient values, but mostly positive. 
India show positive values for the cross-correlation coefficient with Sweden
and both negative and positive values for the coefficient with Italia. For the
other countries, India shows negative cross-correlation for the entire
period visualized, and mostly significantly high values. 
The plot for the cross-correlation between India and Italia is shown as an 
example. The cross-correlation is significantly negative up until a lag of 
-8 days. The value for the coefficient increases, turn positive at a lag
of 3 days and the times series becomes significantly correlated after 13 
days lag. 



### Task C

```{r Principal Component Analysis}

# Selecting columns
df <- select(df, sweden, denmark, norway, britain, italia, india)
# Remove missing values
df <- na.omit(df)
# PCA transform
df.pca <- princomp(df)

# Scores and loadings
# df.pca$scores
# df.pca$loadings

# Summary and plot
# plot(df.pca)
# biplot(df.pca)
summary(df.pca)

# Plotting explained variance
fviz_eig(df.pca)

# Plotting scores
fviz_pca_ind(df.pca,
             col.ind = 'cos2', 
             gradient.cols = c('#00AFBB', '#E7B800', '#FC4E07'),
             repel = FALSE  
)

# Plotting loadings
fviz_pca_var(df.pca,
             col.var = 'contrib', 
             gradient.cols = c('#00AFBB', '#E7B800', '#FC4E07'),
             repel = TRUE   
)

# Biplot: scores and loadings
fviz_pca_biplot(df.pca, repel = FALSE,
                col.var = '#1874CD', 
                col.ind = '#66CDAA'  
)

```
The principal components analysis shows that over 50% of the variance in the
data can be explained by 1 component. The plot of the loadings show that
Denmark, Norway, Great Britain and Italy are correlated as they lie in 
the same direction in along the axis for the first principal component. Denmark
is contributing to most of the variance in the first principal component. 
India contributes to a very small amount of the variance in the dataset. 
Sweden can be considered different compared to the other countries and is 
responsible for almost all the variance in the second component. 



### Task D

```{r Smoothing}

# Smoothing time-series
swe <- zoo::rollmean(df$sweden, k=14, fill=NA)
dnk <- zoo::rollmean(df$denmark, k=14, fill=NA)
nor <- zoo::rollmean(df$norway, k=14, fill=NA)
gbr <- zoo::rollmean(df$britain, k=14, fill=NA)
ita <- zoo::rollmean(df$italia, k=14, fill=NA)
ind <- zoo::rollmean(df$india, k=14, fill=NA)

# Creating a new data.frame with smoothed values
dfS <- data.frame(date, swe, dnk, nor, gbr, ita, ind)

# Renaming data.frame
names(dfS) <- c('date', 'Sweden', 'Denmark', 'Norway',
                'Britain', 'Italia', 'India')

# Remove missing values
dfS <- na.omit(dfS)

```

```{r Line plot of smoothed values}

p <- ggplot(dfS, aes(x=date)) + 
  geom_line(aes(y=Sweden), color='#CD950C') + 
  geom_point(aes(y=Sweden), color='#CD950C', size=0.5) + 
  geom_line(aes(y=Denmark), color='#CD3333') + 
  geom_point(aes(y=Denmark), color='#CD3333', size=0.5) +
  geom_line(aes(y=Norway), color='#009ACD') + 
  geom_point(aes(y=Norway), color='#009ACD', size=0.5) +
  geom_point(aes(y=Britain), color='#8A360F', size=0.5) +
  geom_line(aes(y=Britain), color='#8A360F') + 
  geom_line(aes(y=Italia), color='#9BCD9B') + 
  geom_point(aes(y=Italia), color='#9BCD9B', size=0.5) +
  geom_line(aes(y=India), color='#EE7600') + 
  geom_point(aes(y=India), color='#EE7600', size=0.5) +
  labs(x='', y='New cases', color='legend') 
p

```


```{r PCA on smoothed series}

# Selecting columns 
dfS <- select(dfS, Sweden, Denmark, Norway, 
              Britain, Italia, India)

# PCA transform
dfS.pca <- princomp(dfS, scale=TRUE)
# Scores
# dfS.pca$scores
# Loadings
# dfS.pca$loadings

# Summary and plot
# plot(dfS.pca)
# biplot(dfS.pca)
summary(dfS.pca)

# Plotting explained variance
fviz_eig(dfS.pca)


# Plotting scores
fviz_pca_ind(dfS.pca,
             col.ind = 'cos2', 
             gradient.cols = c('#00AFBB', '#E7B800', '#FC4E07'),
             repel = FALSE  
)

# Plotting loadings
fviz_pca_var(dfS.pca,
             col.var = 'contrib', 
             gradient.cols = c('#00AFBB', '#E7B800', '#FC4E07'),
             repel = TRUE   
)

# biplot: scores and loadings
fviz_pca_biplot(dfS.pca, repel = FALSE,
                col.var = '#1874CD', 
                col.ind = '#66CDAA'  
)


```

The time series was smoothed with a kernel size parameter of k=14. 
When performing the principal component analysis, the variance explained by
the first component increased to 70%. The differences between the previously 
similar countries Denmark, Norway, Great Britain and Italia increased when
performing pca on the smoothed series, and Italia shows to be more similar to 
Sweden. 
