---
title: "Ex04"
output: pdf_document
date: "2022-10-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 4 - Stationarity in ARIMA models

### Task A
### Loading packages

```{r Loading packages, message=FALSE}
library(tseries)
library(forecast)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(imputeTS)
```


### Simulation function

Implementation of the formulas given in the assignment.
Inputs to the function are a list containing the two model-parameters,
the name of the model to simulate and a seed.
The function return the list of values for the model specified in the
input, after cutting the first 100 values. 

```{r Simulation function}
ts_sim <- function(params, model, seed){
  set.seed(seed=seed)
  t_max <- 1000
  n <- t_max + 100
  sigma <- 0.1
  e <- rnorm(n=n, mean=0, sd=sigma)
  k <- 2
  ar <- 0
  ma <- 0
  p <- params
  for(t in 1:length(e)){
    if(t <= k){
      ar[t] <- e[t]
      ma[t] <- e[t]
    }
    else{
      ar[t] <- p[1]*ar[t-1] + p[2]*ar[t-2] + e[t]
      ma[t] <- p[1]*e[t-1] + p[2]*e[t-2] + e[t]
    }
  }
  ar <- ar[101:length(ar)]
  ma <- ma[101:length(ma)]
  if(model=='AR'){
    return(ar)
  } else if(model=='MA'){
    return(ma)
  }
}

```


### Simulating time series

```{r Simulations}
# Model 1
model1 <- list()
params <- c(0.6, -0.3)
seeds <- c(2, 7, 9)

for(i in 1:3){
  seed <- seeds[i]
  sim <- ts_sim(params, 'AR', seed)
  model1 <- append(model1, list(sim))
}

# Model 2
model2 <- list()
params <- c(0.8, 0.2)
seeds <- c(2, 7, 9)

for(i in 1:3){
  seed <- seeds[i]
  sim <- ts_sim(params, 'AR', seed)
  model2 <- append(model2, list(sim))
}

# Model 3
model3 <- list()
params <- c(0.6, -0.3)
seeds <- c(2, 7, 9)

for(i in 1:3){
  seed <- seeds[i]
  sim <- ts_sim(params, 'MA', seed)
  model3 <- append(model3, list(sim))
}

# Model 4
model4 <- list()
params <- c(0.8, 0.2)
seeds <- c(2, 7, 9)

for(i in 1:3){
  seed <- seeds[i]
  sim <- ts_sim(params, 'MA', seed)
  model4 <- append(model4, list(sim))
}

```



### Parameter combinations

#### For Autoregressive model (AR(2))
##### Model assumptions: 
restriction to stationary data and the side conditions for the model 
parameter combinations is given as: 

```{=latex}
$$
\begin{tabular}{ll}
| $\phi_{1}$ | , | $\phi_{2}$ | < 1   \\
$(\phi_{1} + \phi_{2})$ < 1 \\
$(\phi_{2} - \phi_{1})$ < 1 \\
\end{tabular}
$$
```


#### For Moving-Average model (MA(2))
##### Model assumptions: 
restriction to stationary data and the side conditions for the model 
parameter combinations is given as: 

```{=latex}
$$
\begin{tabular}{ll}
| $\theta_{1}$ | , | $\theta_{2}$ | < 1   \\
$(\theta_{1} + \theta_{2})$ > -1 \\
$(\theta_{1} - \theta_{2})$ < 1 \\
\end{tabular}
$$
```


Filling in the values for the parameter combination for each of 
the models, we see that Model 1, Model 3 and Model 4 fulfill the
conditions. Model 2 does not fulfill the side conditions. 


### Task B - Plotting time series, ACF, PACF and KPSS test. 

```{r Model 1, out.width='100%', out.height='100%'}
# Model 1
# Creating data.frame
index1 <- c(1:length(model1[[1]]))
df1 <- data.frame(index1, model1[[1]], model1[[2]], model1[[3]])
names(df1) <- c('index', 'simulation1', 'simulation2', 'simulation3')

# Reshaping
df1_long <- gather(df1, simulation, value, 
                   simulation1:simulation3, factor_key=TRUE)

# Subplots of each simulation
ggplot(data = df1_long) +
  geom_line(mapping = aes(x = index, y = value), color = '#66CDAA') +
  labs(x='', y='') + 
  facet_wrap(~ simulation) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank()) +
  ggtitle('Model 1')


# ACF, PACF and KPSS test
p_list1 = list()
for(i in 1:3){
  ts <- model1[[i]]
  # acf(ts)
  # pacf(ts)
  kpss <- kpss.test(ts)
  p <- kpss['p.value']
  print(p)
  p_list1[i] <- p
}

```
```{r Model 2, out.width='100%', out.height='100%'}
# Model 2
# Creating data.frame
index2 <- c(1:length(model2[[1]]))
df2 <- data.frame(index2, model2[[1]], model2[[2]], model2[[3]])
names(df2) <- c('index', 'simulation1', 'simulation2', 'simulation3')

# Reshaping
df2_long <- gather(df2, simulation, value, 
                   simulation1:simulation3, factor_key=TRUE)

# Subplots of each simulation
ggplot(data = df2_long) +
  geom_line(mapping = aes(x = index, y = value), color = '#00B2EE') +
  labs(x='', y='') + 
  facet_wrap(~ simulation) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank()) +
  ggtitle('Model 2')


# ACF, PACF and KPSS test
p_list2 = list()
for(i in 1:3){
  ts <- model2[[i]]
  # acf(ts)
  # pacf(ts)
  kpss <- kpss.test(ts)
  p <- kpss['p.value']
  print(p)
  p_list2[i] <- p
}

```


```{r Model 3, out.width='100%', out.height='100%'}
# Model 3
# Creating data.frame
index3 <- c(1:length(model3[[1]]))
df3 <- data.frame(index3, model3[[1]], model3[[2]], model3[[3]])
names(df3) <- c('index', 'simulation1', 'simulation2', 'simulation3')

# Reshaping
df3_long <- gather(df3, simulation, value, 
                   simulation1:simulation3, factor_key=TRUE)

# Subplots of each simulation
ggplot(data = df3_long) +
  geom_line(mapping = aes(x = index, y = value), color = '#9370DB') +
  labs(x='', y='') + 
  facet_wrap(~ simulation) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank()) +
  ggtitle('Model 3')

# ACF, PACF and KPSS test
p_list3 = list()
for(i in 1:3){
  ts <- model3[[i]]
  # acf(ts)
  # pacf(ts)
  kpss <- kpss.test(ts)
  p <- kpss['p.value']
  print(p)
  p_list3[i] <- p
}

```


```{r Model 4, out.width='100%', out.height='100%'}
# Model 4
# Creating data.frame
index4 <- c(1:length(model4[[1]]))
df4 <- data.frame(index4, model4[[1]], model4[[2]], model4[[3]])
names(df4) <- c('index', 'simulation1', 'simulation2', 'simulation3')

# Reshaping
df4_long <- gather(df4, simulation, value, simulation1:simulation3, factor_key=TRUE)

# Subplots of each simulation
ggplot(data = df4_long) +
  geom_line(mapping = aes(x = index, y = value), color = '#CD9B9B') +
  labs(x='', y='') + 
  facet_wrap(~ simulation) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank()) +
  ggtitle('Model 4')


# ACF, PACF and KPSS test
p_list4 = list()
for(i in 1:3){
  ts <- model4[[i]]
  # acf(ts)
  # pacf(ts)
  kpss <- kpss.test(ts)
  p <- kpss['p.value']
  print(p)
  p_list4[i] <- p
}


```

From the plots of the time series, the ACFs, PACFs and KPSS-test, 
we see that model 1, 3 and 4 are stationary time series. 
All the mentioned models have a p-value greater than 0.05, 
meaning that they are trend stationary.
Model 2 has a clear trend in the ACF and also a p-value less than 0.05, 
meaning it is not trend stationary. 


### Task C - Fitting parameters

```{r Fitting parameters}
# Fit parameters using auto.arima
# Model 1
model1_parameters = list()
for(i in 1:3){
  ar <- auto.arima(model1[[i]])
  coeff <- ar['coef']
  model1_parameters[i] <- coeff
}
# 0.605, -0.292

# Model 2
model2_parameters = list()
for(i in 1:3){
  ar <- auto.arima(model2[[i]])
  coeff <- ar['coef']
  model2_parameters[i] <- coeff
}
# -0.199

# Model 3
model3_parameters = list()
for(i in 1:3){
  ar <- auto.arima(model3[[i]])
  coeff <- ar['coef']
  model3_parameters[i] <- coeff
}
# 0.608, -0.285

# Model 4
model4_parameters = list()
for(i in 1:3){
  ar <- auto.arima(model4[[i]])
  coeff <- ar['coef']
  model4_parameters[i] <- coeff
}
# 0.804, 0.221
```

Although the estimated model parameters vary some for the three
simulations done for each model, the estimated parameters for the 
stationary models 1, 3 and 4, are very close to the actual parameters.
The estimated model parameters for model 2 are very different from 
the actual parameters. 


\newpage

### Task D - First-order differences

For the AR(2) model, the process is described with the equation:

```{=latex}
$$
x_{t} = \phi_{1} x_{t-1} + \phi_{2} x_{t-2} + \epsilon_{t}
$$
```


By using matrix notation the AR(2) expression can be rewritten 
a first-order difference equation: 

```{=latex}
$$
X_{t} = 
\begin{bmatrix}
y_{t} \\
y_{t-1} \\
\end{bmatrix}
$$
```


Applying the notation to the error term, we get a (2x1) vector:

```{=latex}
$$
v_{t} =
\begin{bmatrix}
\epsilon_{t} \\
0 \\
\end{bmatrix}
$$
```


The coefficients will be contained in a (2x2) matrix, given as:

```{=latex}
$$
M =
\begin{bmatrix}
\phi_{1} & \phi_{2} \\
1 & 0 \\
\end{bmatrix}
$$
```


First-order difference equation: 
```{=latex}
$$
X_{t} = M \ X_{t-1} +  v{t}
$$
```


Which would be equivalent to: 
```{=latex}
$$
\begin{bmatrix}
y_{t} \\
y_{t-1} \\
\end{bmatrix}
=
\begin{bmatrix}
\phi_{1} & \phi_{2} \\
1 & 0 \\
\end{bmatrix}
\begin{bmatrix}
y_{t-1} \\
y_{t-2} \\
\end{bmatrix}
+
\begin{bmatrix}
\epsilon_{t} \\
0 \\
\end{bmatrix}
$$
```


We can use the matrix M to check for stationarity by computing the
eigenvalues. The eigenvalues of M are two solutions for the x
polynomial that satisfy the characteristic equation: 
```{=latex}
$$
x^{t} - \phi_{1} x - \phi_{2} = 0
$$
```


```{=latex}
The eigenvalues, $\lambda_{1}$ and $\lambda_{2}$ must then satisfy 
$(x - \lambda_{1}) \ (x -\lambda_{2})$ and can be found using equation: 
```



```{=latex}

$$
\lambda_{1}, \lambda_{2} = \frac{\phi_{1} \pm \sqrt{\phi_{2}^2 + 4 \phi_{2}} }{2}
$$

```

Stationarity requires that the eigenvalues are less than the absolute value
of 1. For AR(2), this is the case if: 

```{=latex}
$$
\begin{tabular}{ll}
$(\phi_{1} + \phi_{2})$ < 1 \\
$(\phi_{2} - \phi_{1})$ < 1 \\
$ \phi_{2} > -1$
\end{tabular}
$$
```

If the eigenvalues are less than unity, then the system will be stable
and stationary. 
