---
title: "Exercise 1"
output: pdf_document
date: "2022-11-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Loading packages

```{r Loading packages, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(imputeTS)
library(depmixS4)
library(caret)
```


## Exercise 1 - Hidden Markov Models

### Task A

Load the dataset and perform exploratory analysis

```{r Loading data}
# Loading data
jogging <- read.csv('jogging.csv')
column_names <- colnames(jogging)

```


```{r Missing values}
# Missing values
statsNA(jogging$HR)
statsNA(jogging$Speed)
statsNA(jogging$GT)

```


```{r Summary}
# Summary
summary(jogging)

# Unique values for GT
unique(jogging$GT)

```



```{r Plotting1, out.width='75%', out.height='75%'}
# Plotting
plot_hr <- ggplot(jogging, aes(x=Time, y=HR, color=GT)) +
  geom_line()

plot(plot_hr)

```

```{r Plotting2, out.width='75%', out.height='75%'}
# Plotting
plot_speed <- ggplot(jogging, aes(x=Time, y=Speed, color=GT)) +
  geom_line()

plot(plot_speed)

```


From HR plot: Hard to point out specific time points suggesting a 
transition from one underlying state to the next, but it can be
considered that the local minimum- and maximum values in the graph
represent different underlying states. 
From Speed plot: Clustering of measurements around what could be 
considered different underlying states. Change points would in this case
be points where the value either rapidly increases or decreases. 


### Task B


Fitting two Gaussian Hidden Markov Models with 3 latent states.

HR Model

```{r HR Model}
# HR model
set.seed(1)
modelHR <- depmix(HR ~ 1, data=jogging, nstates=3,
                  family=gaussian())

# Parameter estimation
modelHR <- depmixS4::fit(modelHR)

# Summary of model
summary(modelHR)

# Estimated states and probabilities
stateHR <- posterior(modelHR, type='viterbi')

# Adding estimated states to dataframe
jogging$modelHR <- stateHR$state


```

```{r Plotting4, out.width='75%', out.height='75%'}
# Plotting
hr <- ggplot(jogging, aes(x=Time, y=HR, color=modelHR)) +
  geom_line()
plot(hr)

```


Speed model

```{r Speed Model}
# Speed model
set.seed(1)
modelSpeed <- depmix(Speed ~ 1, data=jogging, nstates=3,
                     familiy=gaussian())

# Parameter estimation
modelSpeed <- depmixS4::fit(modelSpeed)

# Summary of model
summary(modelSpeed)

# Estimated states and probabilities
stateSpeed <- posterior(modelSpeed, type='viterbi')

# Adding estimated states to dataframe
jogging$modelSpeed <- stateSpeed$state

```

```{r Plotting5, out.width='75%', out.height='75%'}
# Plotting
speed <- ggplot(jogging, aes(x=Time, y=Speed, color=modelSpeed)) +
  geom_line()
plot(speed)

```


### Task C

Comparing estimated states and ground truth state using a confusion matrix

```{r}
actual = factor(jogging$GT)
predHR = factor(jogging$modelHR)
predSpeed = factor(jogging$modelSpeed)

# Confusion matrix HR
matrixHR <- confusionMatrix(data=predHR, reference=actual)
matrixHR

```

```{r}
# Confusion matrix Speed
matrixSpeed <- confusionMatrix(data=predSpeed, reference=actual)
matrixSpeed

```


### Task D

We see from the plots that the states have been permuted. 
For the HR model all the states have been permuted and for the 
speed model state 2 and state 3 have been permuted. The permutations 
affect the statistics obtained from the confusion matrix. For
the speed model state 1 is not permuted and the confusion matrix shows 
a sensitivity of 0.99.


Assumptions for Hidden Markov Model
```{=latex}
$$
\begin{tabular}{ll}
$X_{t}$ : A Markov chain and not directly observable \\
$Y_{t}$ : Conditionally independent of $X_{t-1}, .., X_{1}$ given $X_{t}$ 
\end{tabular}
$$
```


The models do not fulfill the assumptions, as the observed state is not
independent of the previous observations. 