---
title: "Exercise 1"
output: pdf_document
date: "2022-11-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Loading packages

```{r Loading packages, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(reshape2)
library(tseries)
library(forecast)
library(tidyr)
library(imputeTS)
library(depmixS4)
library(caret)
```


## Exercise 1 - Hidden Markov Models

### Task A

Load the dataset and perform exploratory analysis

```{r Loading data}
# Loading data
jogging <- read.csv('jogging.csv')
column_names <- colnames(jogging)
ts <- ts(jogging)

```


```{r Missing values}
# Missing values
statsNA(ts[, 'HR'])
statsNA(ts[, 'Speed'])
statsNA(ts[, 'GT'])

```


```{r Summary}
# Summary
summary(ts)

```



```{r Plotting1, out.width='100%', out.height='100%'}
# Plotting HR
colors <- c('HR'='#6495ED')
ggplot(jogging, aes(x=Time)) +
  geom_point(aes(y=HR, color='HR'), size=0.5) +
  labs(x='Time', y='Value', color='Measurement') +
  scale_color_manual(values=colors)

```


```{r Plotting2, out.width='100%', out.height='100%'}
# Plotting speed
colors <- c('Speed'='#20B2AA')
ggplot(jogging, aes(x=Time)) +
  geom_point(aes(y=Speed, color='Speed'), size=0.5) +
  labs(x='Time', y='Value', color='Measurement') +
  scale_color_manual(values=colors)

```


```{r Plotting3, out.width='100%', out.height='100%'}
# Speed and GT
colors <- c('Speed'='#20B2AA', 'GT'='#8968CD')
ggplot(jogging, aes(x=Time)) +
  geom_point(aes(y=Speed, color='Speed'), size=0.5) +
  geom_point(aes(y=GT, color='GT'), size=0.5) +
  labs(x='Time', y='Value', color='Measurement') +
  scale_color_manual(values=colors)

```

Any obvious change points, suggesting a transition from one 
underlying state to the next?

From HR plot: Hard to point out specific time points, but it can be
considered that the local minimum- and maximum values in the graph
represent different underlying states. 
From Speed plot: Clustering of measurements around what could be 
considered different underlying states. Change points would in this case
be points where the value either rapidly increases or decreases. 


### Task B


Fitting two Gaussian Hidden Markov Models with 3 latent states.

HR Model

```{r HR Model}
# HR model
set.seed(1)
modelHR <- depmix(HR ~ 1, data=jogging, nstates=3,
                  family=gaussian())

# Parameter estimation
modelHR <- depmixS4::fit(modelHR)

# Summary of model
summary(modelHR)

# Estimated states and probabilities
posteriorHR <- posterior(modelHR, type='viterbi')


```

```{r Plotting4, out.width='75%', out.height='75%'}
# Plotting
# States
ggplot(posteriorHR, aes(x=as.numeric(row.names(posteriorHR)), 
                        y=state, colour = factor(state))) +
  geom_point(size=1.5) + 
  labs(x='Time', y='Value', color='State')

```

```{r, out.width='75%', out.height='75%'}
# Probabilities
colors <- c('S1'='#6495ED', 'S2'='#20B2AA', 'S3'='#8968CD')
ggplot(posteriorHR, aes(x = as.numeric(row.names(posteriorHR)))) +
  geom_point(aes(y=S1, color='S1'), size=1.5) +
  geom_point(aes(y=S2, color='S2'), size=1.5) +
  geom_point(aes(y=S3, color='S3'), size=1.5) +
  labs(x='Time', y='Value', color='StateProb') +
  scale_color_manual(values=colors)
```

Speed model

```{r Speed Model}
# Speed model
set.seed(1)
modelSpeed <- depmix(Speed ~ 1, data=jogging, nstates=3,
                     familiy=gaussian())

# Parameter estimation
modelSpeed <- depmixS4::fit(modelSpeed)

# Summary of model
summary(modelSpeed)

# Estimated states and probabilities
posteriorSpeed <- posterior(modelSpeed, type='viterbi')

```

```{r Plotting5, out.width='75%', out.height='75%'}
# Plotting
# States
ggplot(posteriorSpeed, aes(x=as.numeric(row.names(posteriorSpeed)), 
                        y=state, colour=factor(state))) +
  geom_point(size=1.5) + 
  labs(x='Time', y='Value', color='State')

```

```{r, out.width='75%', out.height='75%'}
# Probabilities
colors <- c('S1'='#6495ED', 'S2'='#20B2AA', 'S3'='#8968CD')
ggplot(posteriorSpeed, aes(x = as.numeric(row.names(posteriorSpeed)))) +
  geom_point(aes(y=S1, color='S1'), size=1.5) +
  geom_point(aes(y=S2, color='S2'), size=1.5) +
  geom_point(aes(y=S3, color='S3'), size=1.5) +
  labs(x='Time', y='Value', color='StateProb') +
  scale_color_manual(values=colors)

```

### Task C

Comparing estimated states and ground truth state using a confusion matrix

```{r}
actual = factor(jogging$GT)
estimated1 = factor(posteriorHR$state)
estimated2 = factor(posteriorSpeed$state)

# Confusion matrix HR model
matrixHR <- confusionMatrix(data=estimated1, reference=actual)
matrixHR

```

```{r}
# Confusion matrix Speed model
matrixSpeed <- confusionMatrix(data=estimated2, reference=actual)
matrixSpeed

```

\newpage


Interpret the results. 
Are the model assumptions for Hidden Markov Models fulfilled in the two scenarios? 


For the HR model: Best at estimating correctly for state 2, 
and 0 Specificity for state 3. 
Speed model: Lower overall accuracy compared to the HR model. 
Almost all the state 1 cases is correctly classified, 
but much lower correct estimations for the other states. 


Assumptions for Hidden Markov Model
```{=latex}
$$
\begin{tabular}{ll}
$X_{t}$ : A Markov chain and not directly observable \\
$Y_{t}$ : Conditionally independent of $X_{t-1}, .., X_{1}$ given $X_{t}$ \\
$Y_{t}$ : Dependent on $X_{t}$ with emission probabilities B
\end{tabular}
$$
```

