---
title: "Exercise 1"
output: pdf_document
date: "2022-11-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Loading packages

```{r Loading packages, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(reshape2)
library(tseries)
library(forecast)
library(tidyr)
library(imputeTS)
library(depmixS4)
library(caret)
```


## Exercise 1 - Hidden Markov Models

### Task A

Load the dataset and perform exploratory analysis

```{r Loading data}
# Loading data
jogging <- read.csv('jogging.csv')
column_names <- colnames(jogging)

```


```{r Missing values}
# Missing values
statsNA(jogging$HR)
statsNA(jogging$Speed)
statsNA(jogging$GT)

```


```{r Summary}
# Summary
summary(jogging)

# Unique values for GT
unique(jogging$GT)

```



```{r Plotting1, out.width='75%', out.height='75%'}
# Plotting
plot_hr <- ggplot(jogging, aes(x=Time, y=HR, color=GT)) +
  geom_line()

plot(plot_hr)

```

```{r Plotting2, out.width='75%', out.height='75%'}
# Plotting
plot_speed <- ggplot(jogging, aes(x=Time, y=Speed, color=GT)) +
  geom_line()

plot(plot_speed)

```


Any obvious change points, suggesting a transition from one 
underlying state to the next?

From HR plot: Hard to point out specific time points, but it can be
considered that the local minimum- and maximum values in the graph
represent different underlying states. 
From Speed plot: Clustering of measurements around what could be 
considered different underlying states. Change points would in this case
be points where the value either rapidly increases or decreases. 


### Task B


Fitting two Gaussian Hidden Markov Models with 3 latent states.

HR Model

```{r HR Model}
# HR model
set.seed(1)
modelHR <- depmix(HR ~ 1, data=jogging, nstates=3,
                  family=gaussian())

# Parameter estimation
modelHR <- depmixS4::fit(modelHR)

# Summary of model
summary(modelHR)

# Estimated states and probabilities
stateHR <- posterior(modelHR, type='viterbi')

# Adding estimated states to dataframe
jogging$modelHR <- stateHR$state


```

```{r Plotting4, out.width='75%', out.height='75%'}
# Plotting
hr <- ggplot(jogging, aes(x=Time, y=HR, color=modelHR)) +
  geom_line()
plot(hr)

```


Speed model

```{r Speed Model}
# Speed model
set.seed(1)
modelSpeed <- depmix(Speed ~ 1, data=jogging, nstates=3,
                     familiy=gaussian())

# Parameter estimation
modelSpeed <- depmixS4::fit(modelSpeed)

# Summary of model
summary(modelSpeed)

# Estimated states and probabilities
stateSpeed <- posterior(modelSpeed, type='viterbi')

# Adding estimated states to dataframe
jogging$modelSpeed <- stateSpeed$state

```

```{r Plotting5, out.width='75%', out.height='75%'}
# Plotting
speed <- ggplot(jogging, aes(x=Time, y=Speed, color=modelSpeed)) +
  geom_line()
plot(speed)

```


### Task C

Comparing estimated states and ground truth state using a confusion matrix

```{r}
actual = factor(jogging$GT)
predictionHR = factor(jogging$modelHR)
predictionSpeed = factor(jogging$modelSpeed)

# Confusion matrix HR
matrixHR <- confusionMatrix(data=predictionHR, reference=actual)
matrixHR

```

```{r}
# Confusion matrix Speed
matrixSpeed <- confusionMatrix(data=predictionSpeed, reference=actual)
matrixSpeed

```

\newpage


Interpret the results. 
Are the model assumptions for Hidden Markov Models fulfilled in the two scenarios? 


For the HR model: Best at estimating correctly for state 2, 
and 0 Specificity for state 3. 
Speed model: Lower overall accuracy compared to the HR model. 
Almost all the state 1 cases is correctly classified, 
but much lower correct estimations for the other states. 


Assumptions for Hidden Markov Model
```{=latex}
$$
\begin{tabular}{ll}
$X_{t}$ : A Markov chain and not directly observable \\
$Y_{t}$ : Conditionally independent of $X_{t-1}, .., X_{1}$ given $X_{t}$ \\
$Y_{t}$ : Dependent on $X_{t}$ with emission probabilities B
\end{tabular}
$$
```

